Psilocin fosters neuroplasticity in iPSC-derived human cortical neurons

Version 1

Abstract

Abstract

Psilocybin is studied as innovative medication in anxiety, substance abuse and treatment-resistant depression. Animal studies show that psychedelics promote neuronal plasticity by strengthening synaptic responses and protein synthesis. However, the exact molecular and cellular changes induced by psilocybin in the human brain are not known. Here, we treated human cortical neurons derived from induced pluripotent stem cells with the 5-HT2A receptor agonist psilocin - the psychoactive metabolite of psilocybin. We analyzed how exposure to psilocin affects 5-HT2A receptor localization, gene expression, neuronal morphology, synaptic markers and neuronal function. Upon exposure of human neurons to psilocin, we observed a decrease of cell surface-located 5-HT2A receptors first in the axonal-followed by the somatodendritic-compartment. Psilocin further provoked a 5-HT2A-R-mediated augmentation of BDNF abundance. Transcriptomic profiling identified gene expression signatures priming neurons to neuroplasticity. On a morphological level, psilocin induced enhanced neuronal complexity and increased expression of synaptic proteins, in particular in the postsynaptic-compartment. Consistently, we observed an increased excitability and enhanced synaptic network activity in neurons treated with psilocin. In conclusion, exposure of human neurons to psilocin might induces a state of enhanced neuronal plasticity which could explain why psilocin is beneficial in the treatment of neuropsychiatrie disorders where synaptic dysfunctions are discussed.

Introduction

Since the 90’s the mind-altering psychedelics have become again the subject of an intriguing stream of research in psychiatric disorders1,2. Particularly, in times where the development and the approval of new medications is decreasing and the number of patients suffering from psychiatric disorders is rising there is a huge need for new therapeutical interventions3,4. Compared to classical drugs the broad therapeutical effect of psychedelics is rapid, robust and can be long-lasting even after single administration5. In particular, the 5-hydroxytryptamine receptor 2A (5-HT2A-R) targeting psilocybin, the compound of the so-called “magic mushrooms”, is discussed as fast-acting and long-lasting antidepressant in treatment-resistant depression (TRD), anxiety, obsessive-compulsive disorder and addiction6–11. Not surprisingly psilocybin is stated as “breakthrough therapy” by the United States Food and Drug Administration (FDA) in the treatment of depression since 2019. While we know that 5-HT2A-R activation plays a principal role in serotonergic psychedelic-mediated behavioral and cellular response12–22 the molecular and cellular changes induced in the brain - on a single cell and network level are barley understood. The recent demonstration of intracellular 5-HT2A-Rs additionally increased the complexity of psychedelic actions23. The administration of psychedelics may enable brain network resetting1 by generating a plastic cellular state in which synaptic remodeling and augmentation of neuroplasticity-associated proteins and genes are likely24. Indeed, biological evidence for the “resetting” hypothesis comes from a pioneering study by Olson and colleagues in 2018 which showed that the treatment with serotonergic psychedelics increases the synthesis of synaptic proteins, strengthens synaptic responses, and fosters neurito- and synaptogenesis in rat cortical neurons25. The group therefore introduced the term “psychoplastogen” (greek: psych- (mind), - plast (molded), - gen (producing) for underlining their plasticity-promoting properties. Moreover, serotonergic psychedelics already have been shown to promote the 5-HT2A-R-mediated growth of dendritic spines and modulate neurotransmission25–27. As molecular and cellular psychedelic research is recently based nearly exclusively on animal studies, the question emerged whether those insights can be translated to the human brain. Psychiatric disorders and psychedelic effects are complex, multisymptomatic and therefore often difficult to study in non-human model organisms. Most importantly, drugs that are efficiently tested in psychiatric animal models might not be necessarily transferable to the human system28. In that context, induced human pluripotent stem cells (IPSC) have emerged as a powerful tool to generate neurons and neuronal circuitries, model brain disorders and identity the molecular mechanisms of drug interventions29.

Here, we explored the molecular, transcriptional, morphometric and functional consequences of the psychoactive 5-HT2A receptor agonist psilocin, the active metabolite of psilocybin in human iPSC-derived cortical neurons. We demonstrate that psilocin leads to a set of molecular, morphological and functional changes that start shortly after administration and manifest in time. These include an increase in brain derived neurotrophic factor (BDNF) expression and an activation of gene expression programs associated with neuromodulation and plasticity resulting in increased neuronal complexity, synaptogenesis and changes in neuronal network function. Thus, our study provides first evidence that the 5TH2A receptor agonist psilocybin activates widespread neuroplastic programs in human neurons.

Results

Psilocin causes 5-HT2A receptor internalization and redistribution in human cortical neurons

As an experimental model to study the effects of psilocin in human neurons and neuronal networks we differentiated human iPSCs (expressing the pluripotency-associated transcription factors 0CT4 and S0X2) into neural progenitor cells which we further differentiated into neurons of mainly dorsal forebrain identity (Fig. 1A and Fig. S1A-B)30–33. Neural progenitor cells express typical neural stem I progenitor cell markers such as NESTIN or S0X2 as well as the dorsal forebrain-associated transcription factors PAX6 and are negative for F0XA2, as floorplate/ midbrain marker (Fig. S1B). Neurons differentiated from these progenitors for >5 weeks express the neuronal antigen NeuN, the dendritic marker MAP2, and the axonal marker TAU. Most neurons express TBR1 and I or CTIP2, transcription factors typically found in Layer 5/6 neurons of the human cortex (Fig. 1A). Neuronal activity in mature neurons is indicated by expression of the activity-dependent immediate early gene cFOS (Fig. 1A). Astrocytes expressing glial fibrillary acid protein (GFAP) were found occasionally (Fig. 1A). These represent the only contamination in the otherwise purely neuronal culture, which contains no residual progenitors (Fig. S1C). 5-HT2A-R expression was confirmed by PCR (Fig. S1D) and by immunocytochemistry targeting an extracellular epitope which showed presentation of the receptor in the somatodendritic and axonal compartment (Fig. 1B, upper row). 5-HT2A-R agonists such as psilocin are hypothesized to cause immediate (within minutes) internalization of the 5-HT2A-R in vivo and in vitro and are described to cause desensitization by reduced receptor abundance34,35. We thus analyzed the effects of psilocin on the 5-HT2A-R in human neurons 10 minutes (min) and 24 hours (hrs) following exposure to psilocin. By 10 min, 5-HT2A-R expression at the neuronal surface significantly decreased in the axonal compartment (Tau-positive; from MCtrl = 0.4 ± 0.3; M10min = 0.3 ± 0.3), but not in the somatodendritic compartment (MAP2-positive; measured as 5-HT2A-R particles per μm; Fig. 1B-C, F).

Validation of mature cortical neuron properties and effect of psilocin on cell surface-located 5-HT2A receptor presentation.

(A) hiPSCs were differentiated in-vitro into glutamatergic cortical neurons over a neuronal progenitor step since the cerebral cortex is a key region for psychedelic effects and psychiatric disorders. Mature 40-day old cortical neurons were glutamatergic (vGLUT2 expression) and expressed cortical layer markers like TBR1, CTIP2 and the neuronal markers NeuN, MAP2 and TAU. cFOS expression is a sign for neuronal activity. GFAP expression indicated a low amount of astrocytes. Scale bar: 50 μm, for GFAP staining 100 μm. (B) Cortical neurons expressed the cell surface-located 5-HT2A receptor (ex5-HT2A-R) validated by immunofluorescence. (B – C) Representative ex5-HT2A-R staining showed decrease in TAU+ axonal receptor presentation 10 minutes after 10 μM psilocin trigger (10 min Psi) compared to the untreated condition (Ctrl), that may indicate receptor complex formation or internalization. (B) Scale bar: 50 μm (C), close up scale bar: 10 μm. (D) Representative ex5-HT2A-R staining. After 24 hrs, both the axonal and dendritic receptor density was significantly decreased, speaking for a subacute down-regulation of the receptor, scale bar: 50 μm (E) close up scale bar: 5 μm. (F) Decrease in TALT axonal receptor localization 10 minutes after 10 μM psilocin trigger compared to the control condition was significant. After 24 hrs, both the axonal and dendritic receptor density was significantly decreased, speaking for a down-regulation of the receptor. For Ctrl N = 180 neurites, for 10 min N = 180 neurites, for 24 hrs N = 135 – 140 neurites were calculated. Two control cell lines each with two biological batches for MAP2+ and TAU+ ex5-HT2A-R were included (except of 24hrs for Ctrl 3 with one biological batch). Kruskal-Wallis-Test for independent samples was calculated. Significance levels against the respective control and for multiple group comparisons. Bonferroni-correction, adjusted p <.05, mean ± SD. (G) Graphic illustration of data in (F), generated with BioRender.com.

The decrease in cell surface-located receptor signals was largely prevented by inhibiting clathrin-mediated endocytosis (CME) using dynamin inhibitor dynasore, indicating that this acute down regulation is at least in part mediated by receptor internalization. Dynasore-mediated internalization of 5-HT2A-R resulted in accumulation of receptor-dependent fluorescence signals in roundish aggregates along the axonal compartment (Fig. S1E-F). At 24 hrs, both, receptor densities on axons (from MCtrl = 0.4 ± 0.3; M24hrs = 0.1 ± 0.2) and dendrites (from MCtrl = 0.6 ± 0.4; M24hrs = 0.2 ± 0.2) were significantly decreased (Fig. 1D-E, F), suggesting a subacute downregulation and/or internalization of the receptor also in the somatodendritic compartment. Thus, psilocin treatment induces a two-phasic decrease of cell-surface located 5-HTR2A-R abundance, first in axons and then in the somatodendritic compartment (Fig. 1G).

Psilocin induces BDNF expression and downstream signaling in human neurons

Psychedelics have been shown to efficiently increase levels of neurotrophic factors, such as BDNF12,25,36,37. We thus addressed BDNF expression in human neurons following exposure to psilocin. We analyzed the effect of different psilocin concentrations (ranging from 10 nM to 10μM) exposed for 10 min and also at 24 hrs following exposure. We observed a dosage-dependent increase in the abundance of BDNF-positive particles (quantified as particles per μm neurite length) which reached significance at a concentration of 10 μM psilocin (Fig. S2A-B). A longer treatment (6 hrs instead of 10 min) with a lower concentration (100 nM instead of 10 μM) of psilocin also induced BDNF abundance significantly, but less strikingly (Fig. S2C-D). We thus continued to use 10 μM psilocin in the following experiments. The clear induction of BDNF particle density was reproduced in four biological independent control cell lines and several differentiation batches for each line (from MCtrl = 1.5 ± 1.2; MPsi = 2.6 ± 1.9) (Fig. 2A-C). When comparing BDNF densities independently in dendrites (MAP2-positive) and axons (Tau-positive), the significant increase in BDNF could be assigned to both neuronal compartments to a comparable extent (Fig. S2E-F). Also, at 48 hrs following the exposure to psilocin, a significant increase of BDNF particles was observed compared to the untreated condition (Fig. S2G). The increase in BDNF abundance was 5-HT2A-R-mediated as treatment of the neurons with the specific 5-HTR2A-R antagonist ketanserin prevented BDNF upregulation by psilocin (Fig. 2D-E) (from MCtrl = 2.5 ± 1.4; MPsi = 4 ± 2.3; MPsi + Ket = 2.1 ±1.4; MKet = 1.4 ± 1.2). Upregulation of BDNF could also be prevented by blocking CME with dynasore or protein kinase C (PKC) with the PKC inhibitor chelerythrine, indicating that CME and PKC activation are critically involved in this process (Fig. 2F-G) (from MCtrl= 3.1 ± 1.5 to MPsi = 4.6 ± 2.4; MPsi + Chel = 3 ± 1.7; MPsi + D = 3.4 ± 1). Upregulation of BDNF was also validated by Western immunoblot where we observed an increase in the levels of pro-BDNF and were able to detect mature BDNF at approximately 14 kDa, which, under baseline conditions, was below the detection level (Fig. S2H). Increased BDNF signaling in psilocin-treated neurons is indicated by an increase in the phosphorylation of the BDNF receptor TrkB (Fig. S2H). Furthermore, as a downstream target of BDNF signaling we observed an increase of the phosphorylation of AKT at Ser 473 1 day (Fig. S2I) and 3 days after psilocin exposure (Fig. 2H-I) which was reversed by ketanserin treatment (from MCtrl = 0.8 ± 0.2; MPsi = 1.3 ± 0.2; MKet = 0.9 ± 0.2).

Psilocin-induced increase in BDNF level was 5-HT2A receptor and PKC- and endocytosis mediated and induced activation of m-BDNF/TrkB-associated downstream pathway.

(A-B) Representative image of a neuronal network for pre-treatment condition (Ctrl) and 24 hrs after a 10 min 10 μM short psilocin trigger (Psi). Scale bar: 50 μm, (B) close up: 2 μm. (C) BDNF density was significantly increased 24 hrs after a 10 min 10 μM psilocin trigger (Psi). Four Ctrl cell lines were included in the analysis (Ctrl with A/ = 613 neuntes; Psi with N = 529 neuntes). (D) Representative image 24 hrs after the simultaneous treatment with psilocin and ketanserin (Psi + Ket) and single treatment with ketanserin (Ket), scale bar: 50 μm. (E) Ketanserin co-treatment (Psi + Ket) and ketanserin monotreatment (Ket) significantly reduced BDNF density compared to the 24 hrs monotreatment psilocin condition, suggesting a 5-HT2A-mediated process. Ketanserin monotreatment provoked a significant reduction in BDNF density compared to ketanserin co-treatment with psilocin. Three Ctrl cell lines were included in the analysis, each with one biological batch (Ctrl with N = 144 neurites; Psi with N = 147 neurites; Psi + Ket with N = 147 neurites; Ket with N = 102 neurites). (F) Representative image 24 hrs after the simultaneous treatment of 10 μM psilocin with chelerythrine (Psi + Chel) or dynasore (Psi + D), scale bar: 50 μm. (G) Chelerythrine (Psi + Chel, selective PKC inhibitor) or dynasore (Psi + D, inhibition of clathrin-coated vesicle invagination) co-treatment significantly reduced BDNF density compared to the psilocin monotreatment condition (Psi). Two Ctrl cell lines were included in the analysis, each with one biological batch. (Ctrl with N = 90 neurites; Psi with N = 90 neurites; Psi + Chel with N = 90; Psi + D with N = 90 neurites). (H) Phosphorylated AKT (pAKT) protein level was increased 72 hrs after psilocin exposure (Psi), reversed by ketanserin co-treatment (Psi + Ket), (I) both effects were not significant. Ctrl cell line 3 was included in the analysis (Ctrl with N = 4 data points, Psi with N = 4 data points, Psi + Ket with N = 4 data points), each with two biological batches. For all analyses Kruskal-Wallis-Test for independent samples was calculated. Post hoc Wilcoxon rank sum test. Bonferroni-correction, adjusted p < .05, mean ± SD. Significance levels against the respective control and for multiple group comparisons are *p <.05.

Psilocin induces gene expression changes associated with axonal and synaptic plasticity

To analyze the influence of psilocin on global gene expression in cortical neurons, we performed whole transcriptome sequencing one day and three days following a single 10 min administration with 10 μM psilocin in neurons from two independent genetic backgrounds. GO enrichment and KEGG pathway analysis comparing cells one day and three days after psilocin administration with the respective controls revealed an enrichment of significantly affected genes in many ontologies and pathways associated with axonal growth and synaptic remodeling, plasticity and learning, memory and cognition (Fig. 3A, S3A). Most significant changes occurred within the first 24 hrs which is reflected when plotting the significantly changed genes only for selected GO term enriched at both time points (Fig. 3B, S3B). When looking at a more global gene expression level of all genes included in the respective GOs, gene expression of GOs associated with axonal outgrowth show a generally higher expression at day 1 which is further increased at day 3. In contrast genes associated to synaptic organization, plasticity and learning, memory and cognition show a trend towards decreased expression at day 1 and a generally stronger increase at day 3 (Fig. 3C, S3C). The overlapping assignment of significant differentially expressed genes to the aforementioned GOs underlines their importance in long-term plasticity (e.g., CDK5, CAMK), synapse formation (e.g., SYN1) and axonal and neurite growth (e.g., GAP43) and neuronal structure marker (e.g., MAPT) (Fig. 3D). The latter constent with increasing staining intensity for TAU (Fig. 2). We further found a strong significant upregulation of immediate early genes (lEGs) (e.g., FOSL2, JUND, EGR1, ARC, FOSB) and of the excitatory glutamatergic AMPA/NMDA receptor genes (GRIA/GRIN) after psilocin administration (Fig. 3E). One day but not three days of psilocin treatment more stongly upregulated the AMPA genes GRIA1,2,3 than the NMDA genes GRIN1,2B,2C (Fig. 3E), supporting modulation of glutamatergic excitatory signalling genes upon psilocin treatment. These effects could be reversed upon co-treatment with antagonist ketanserin, indicating that most changes are 5-HT2A receptor mediated (Fig. 3F).

Psilocin displays fast and enduring changes of the genetic landscape.

(A) Enrichment in differentially expressed genes associated with synapse formation, neuronal plasticity and axonogenesis GO terms 1 day after psilocin administration, and the effect 3 days later. (B) Psilocin induced within 24 hours a first wave of upregulation of selected GO genes based on DESeq2 normalized counts. Log2fold changes of each differentially expressed genes between two conditions are shown as dots. Zscores indicate if more genes in the respective GO term are upregulated or downregulated also indicated by height and color of the bars (C) TPM-normalized mean (red line) of psilocin-induced temporal expression pattern of genes belonging to the indicated GO. Shaded area is indicating SD. (D) Chord plots showing significant genes appearing in at least 4 GO terms (of 6 selected GO terms) 1 day after psilocin administration and appearing in at least 3 GO terms 3 days after psilocin administration. (E). Heat map (z-scaled normalized counts) showing expression of immediate early genes and AMPA/ NMDA receptor genes throughout psilocin administration. (F) Differentially expressed genes, that are up-/ downregulated upon psilocin treatment, showed a reversed effect upon co-treatment with ketanserin. Mean TPM-normalized expression values are shown (red line), shaded area indicates S.D. BP, biological process; CO, Cellular Compartment; KEGG, Kyoto Encyclopedia of Genes and Genomes. TPM, transcripts per kilobase million. Significance levels against the respective control ns: p-adj. >.05, *: p-adj.<=.05, **: p-adj.<=.01, ***: p-adj.<=0.001.

Psilocin increases neunte complexity

The gene expression analysis indicated an induction of morphometric changes by psilocin in particular on neurites. Therefore, we assessed neuronal complexity performing Sholl analysis in neurons differentiated for >40 days at 24 hrs and 48 hrs following psilocin exposure. To identify the dendritic arbor of single mature neurons in these cultures, the cultures were transduced with adeno-associated vectors (AAVs) coding for mCherry under control of the CaMKIla promoter (Fig. 4A-D).

Psilocin induced neurite branching.

(A-C) Cells were transduced with AAV CamKIla p-hCHR2(134a)-mCherry. (A) Representative image for pre-treatment condition (Ctrl) and 48 hrs after a 10 min short psilocin trigger (48 hrs) for mCherry staining showed an increase in neurite intersections for the latter. (B) Singular analysis: 48 hrs after a 10 min 10 μM psilocin trigger the number of primary (25 μm distance from soma) neurite intersections significantly increased compared to the untreated control condition (Ctrl). The number of intersections at 50 μm distance from soma significantly increased 24 hrs and 48 hrs after a 10 min 10 μM psilocin trigger. (C) Significant changes for the calculated total neurite length and for the total number of neurite intersections significantly increased 24 hrs and 48 hrs after a 10 min 10 μM psilocin trigger. Two Control cell lines with two biological batches (Ctrl with N = 43 – 48 neurites; 24 hrs with N = 47 – 48 neurites; 48 hrs with N = 31 – 35 neurites) were included. (D) Sholl analyses summary for results shown in figure (B). For all analyses Kruskal-Wallis-Test for independent samples was calculated. Post hoc Wilcoxon rank sum test. Bonferroni-correction, adjusted p < .05, mean ± SD. Significance levels against the respective control are *p <.05.

We observed an increase of primary neurites at 25 μm (significant at 48 hrs and a trend at 24 hrs from MCtrl = 4.1 ± 2.1; M24hrs = 5.4 ± 3.7; M48hrs = 5.5 ± 2.4) and of intersections at 50 μm (significantly increased at 24 and 48 hrs post treatment from MCtrl = 2.7 ± 1.4; M24hrs = 4 ± 2.3 to M48hrs = 3.9 ± 1.8; Fig. 4B). As a result, the calculated total neurite length was significantly increased at 24 and 48 hrs post treatment from MCtrl = 272.5 ± 142.7; M24hrs = 383.2 ± 209.9; M48hrs = 352.8 ± 162.5 (Fig. 4C). And we observed a significant increase in the total number of intersections in psilocin-exposed neurons quantified in steps of 25 μm up to a distance of 125 μm (MCtrl = 10.9 ± 5.7; M24hrs = 15.3 ± 8.4, M48hrs = 14.1 ± 6.5; Fig. 4D). These data indicate, that gene expression changes elicited by psilocin resulted in neurite outgrowth and an increase of global dendritic complexity as early as 24h following a single 10 min exposure.

Psilocin increases synaptic strength and synaptogenesis

To address changes on neuronal function, we exposed neuronal cultures 10 min or 24 h to 10 μM psilocin and performed whole-cell patch-clamp experiments one week later. Following the 10 min exposure, we observed a increasing trend in the number of evoked action potentials (APs), having amplitudes of around 100 mV (Fig. 5A). Extended exposure of the cultures to psilocin (24h) increased this effect reaching significance (total number of APs from MCtrl = 20.1 ± 12; MPsi 10min; day 7 = 28.9 ± 16.9; MPsi 24hrs; day 7 = 30 ± 14.5; AP amplitude from MCtrl = 101.1 ± 13.3; MPsi 10min; day 7 = 101.1 ± 13.2; MPsi 24hrs; day 7 = 101.3 ± 16.7). To find out whether the increase in AP firing enhanced synaptic network activity, we recorded spontaneous AP-dependent excitatory postsynaptic currents (sEPSCs). Indeed, the frequency of sEPSCs increased in both psilocin-treated conditions to some extent (MCtrl = 0.6 ± 0.7; MPsi 10min; day 7 = 1.0 ± 1.3; MPsi 24hrs; day 7 = 1.1 ±1.2; Fig. 5B). And, we observed a slight increase in sEPSC amplitude (MCtrl = 14.2 ± 4.7; MPsi 10min; day 7 = 16.7 ± 5.1; MPsi 24hrs; day 7 = 15.5 ± 5.3; Fig. 5B), consistent with the observed GRIA upregulation (Fig. 3E). The latter amplitude increase was next examined on AP-independent miniature EPSCs (mEPSCs) which are generally caused by the spontaneous release of single vesicles (Figs. 5C, S4A). The amplitudes of mEPSCs increased 7 days after the start of 24 hrs psilocin administration consistently for two cell lines (MCtrl = 11.1 ± 2.2; MPsi 24hrs; day 7 = 12.8 ± 4.2, Fig. 5C; MCtrl = 9.3 ± 2.8; MPsi 24hrs; day 7 = 11.2 ± 3.7, Fig. S4A), whereas the frequencies of the mEPSCs were rather stable (MCtrl = 0.8 ± 0.5; MPsi 24hrs; day 7 = 0.9 ± 0.6, Fig. 5C; MCtrl = 0.9 ± 0.7; MPsi 24hrs; day 7 = 0.7 ± 0.5, Fig. S4A). When examining mEPSCs from separately differentiated neurons exposed to psilocin for 24 hrs and analyzed at day 11, the increase in mEPSC amplitude was somewhat attenuated compared with 7 days (MCtrl = 11.7 ± 4.3; MPsi 24hrs; day 11 = 12.4 ± 3.6; frequency from MCtrl = 0.8 ± 0.8; MPsi 24hrs; day 11 =0.5 ± 0.3, Fig. 5C). Repeated LSD administration altered gene and protein expression related to neuroplasticity signaling in the mouse prefrontal cortex and increased dendritic spine density38,39. We therefore hypothesized that extended psilocin administration fosters synaptic network activity and synaptogenesis and thus examined effects of a 96 hrs exposure. In line with the above-mentioned hypothesis, the amplitudes of the sEPSCs increased significantly in two lines (MCtrl = 22.5 ± 6.5; MPsi 96hrs; day 10 = 30.7 ± 4.8, Fig. 5D; MCtrl = 15.9 ± 4.5 ; MPsi 96hrs; day 10 = 27.6 ± 10, Fig. S4B), and the frequencies showed a trend to increase similar to Fig. 5B, reflecting enhanced network activity (MCtrl = 1.0 ± 1.0; MPsi 96hrs; day 10 = 1.5 ± 1.0, Fig. 5E; MCtrl = 1.5 ± 1.7; MPsi 96hrs; day 10 = 1.0 ± 0.7, Fig. S4B).

Psilocin-induced increase in synaptic strength and synaptogenesis.

(A) Total number of evoked action potentials (eAPs) significantly increased at day 7 after 24hrs permanent psilocin administration (Psi 24hrs; day 7) and was increased 7 days after a 10 minutes psilocin trigger (Psi 10min; day 7). AP amplitudes stayed constant. Representative traces for total number of APs. One Ctrl cell line with 2 biological batches was included in the analysis. Ctrl 3 with N = 28 cells, Psi 10min; day 7 with N = 29 cells, Psi 24hrs; day 7 with N = 34 cells. (B) Increase in sEPSCs amplitude and frequency 7 days after 10 minutes and 24hrs permanent psilocin administration. One Ctrl cell line with 2 biological batches was included in the analysis. Ctrl 3 with N = 19 cells, Psi 10min; day 7 with N = 20 cells, Psi 24hrs; day 7 with N = 21 cells. (C) Increase in mEPSCs amplitude 6 days after permanent 10 μM psilocin administration, Ctrl 3 with N = 16 cells, Psi 24hrs; day 7 with N = 14 cells. Increase in mEPSCs amplitude 10 days after 24hrs permanent psilocin administration, Ctrl 3 with N = 15 cells, Psi 24hrs; day 11 with N = 20 cells. (D) Significant increase in sEPSCs amplitude 6 days after 96hrs permanent 10 μM psilocin administration (Psi 96hrs; day 10). Ctrl 3 with N = 18 cells, Psi 96hrs; day 10 with N = 13 cells. Representative traces for control and psilocin condition. For all experiments Mann-Whitney-U-test for independent samples was calculated, Mean ± SEM. Significance levels against the respective control are *p <.05. (E-F) Representative dendritic PSD-95 and synapsin staining for the untreated condition (Ctrl), 4 days after permanent 10 μM psilocin administration (4 days) and 10 days after 4 days permanent 10 μM psilocin administration (10 days). (E) Scale bar: 50 μm (F), closeup: 10 μm. Trend in synapsin (G) density and (H) intensity increase 4 days after permanent psilocin administration. (I) PSD-95 particle per neurite length and (J) intensity per area and (K) synapsin/ PSD-95 co-localization were significantly increased 4 and 10 days after 4 days permanent 10 μM psilocin treatment compared to an untreated control condition. (G-K) Ctrl with N = 180 neurites, 4 days with N = 180 neurites, 10 days with N = 180 neurites. Two control cell lines each with two biological batches were included. For all analyses Kruskal-Wallis-Test for independent samples was calculated. Bonferroni-correction, adjusted p <.05, mean ± SD. Significance levels against the respective control are *p <.05.

This enhanced synaptic network activity should affect the abundance of either the presynaptic marker synapsin and/or the postsynaptic marker PSD-95 and may even affect their co-localization at day 4 and at day 10 following 4 day permanent 10 μM psilocin administration. In line with above electrophysiological results for the 10 day psilocin exposure (Fig. 5D) and the gene expression data, which indicated changes associated with synaptogenesis and synaptic plasticity (Fig. 3) we observed for the pre-synaptic marker synapsin, a trend towards an increase of the density (particles per neurite length) (from MCtrl = 0.25 ± 0.2; M4days = 0.28 ± 0.2; MIOdays = 0.29 ± 0.2) and intensity (mean fluorescence intensity of the particles) (from MCtrl = 21 ± 13.3; M4days = 24.1± 15; MIOdays = 19.8 ±10.9) at both time points (Fig. 5E-H). In line with the gene expression data (Fig. 3A, postsynaptic density), we observed a significant increase in the density of the postsynaptic marker PSD-95 (from MCtrl = 1.1 ± 0.5; M4days = 1.4 ± 0.6; MIOdays = 1.3 ± 0.5) and in its intensity (from MCtrl = 31.8 ± 12,9; M4days = 36.8 ± 15.6; MIOdays = 38 ± 16.4) for both time points (Fig. 5I-J). These effects were accompanied by a significant increase in the co-localization of both markers (from MCtrl = 0.2 ± 0.2; M4days = 0.3 ± 0.2; MIOdays = 0.3 ± 0.2; Fig. 5K). Together, these experiments indicate that psilocin augments synaptic strength primarily via an increase in the postsynaptic receptor density, which lasts at least for 6 days from the start of psilocin withdrawal and that extended exposure of neurons to psilocin pronounces the effects on synaptic strength.

Discussion

The neuroplasticity-promoting psychoplastogen25 psilocybin is currently developed as a new medication in the treatment of psychiatric disorders1,2,40. “Brain network resetting”1 by psilocybin , i.e. the restoration of neuronal and synaptic dysfunction associated with the pathophysiology of mental disorders41–45, could be achieved by stimulating signaling pathways associated with neuroplasticity and BDNF signaling24,41. So far, experimental evidence for this hypothesis is based on data from non-human model organisms. In our experimental setting, human neurons exposed to psilocin present with a biphasic downregulation of cell surface-located 5-HT2A-Rs, which starts in the axonal compartment and expands to the dendritic compartment. Stimulation of 5-HT2A-Rs can lead to different adaptive processes including internalization, downregulation and recycling42. Interestingly, pychedelics have been already shown to reduce 5-HT2A-R density in animal models35,43,44 We see a decrease in receptor surface density, prevented by inhibiting the GTPase dynamin arguing that clathrin-mediated endocytosis is involved in this process, which suggests internalization of cell surface-located 5-HT2A-R. Given the potential of psychedelics to diffuse across cell membranes and to activate intracellular pathways23, enhanced internalization of 5-HT2A-R may result in stronger intracellular signaling processes caused by the combination of retrograde trafficking and the activation of internally residing receptors adding both to serotonin-independent and psychedelic-dependent signaling strength45.

We also observed a pronounced psilocin-induced BDNF upregulation in human cortical neurons and show that PKC activation and endocytosis, two mechanisms contributing to 5-HT2A-R internalization46,47, are involved as chelerythrine and dynasore inhibited BDNF augmentation. BDNF plays an important role in neurogenesis, synaptogenesis, and the formation of synaptic interactions. Particularly the mature BDNF/phosphorylated-TrkB receptor complex is involved in multiple pathways linked to prosurvival and synaptic plasticity48. Recently it was shown that psychedelics also directly bind to the BDNF TrKB receptor, thereby affecting TrkB dimerization and facilitating the effect of endogenous BDNF, which underlines the importance of the BDNF system for the action of psychedelics72. In keeping with this, we show that psilocin leads to a fast and enduring upregulation of proteins and genes linked to neuronal complexity, synaptogenesis and synaptic transmission, as also demonstrated in psychedelic-mediated plasticity in animal studies25,26,49. Of note, BDNF upregulation and changes in gene expression where reversed by ketanserin confirming that these effects are medated by 5-HT2A-Rs.

In our model, augmentation of neuronal complexity is an outcome which can be detected as early as 24 hrs after psilocin administration. In rodents, serotonergic psychedelics also foster synaptogenesis and spinogenesis25,26,49. In line with this we show that psilocin also promotes synaptogenesis, measured by the increase in PSD-95, synapsin and their colocalization. As a result, we observe changes in intrinsic neuronal properties and network function. More specifically, we see an increased excitability and an increase in postsynaptic current frequency but in particular amplitude. The higher number of action potentials generated by current injections could be due to increased dendritic excitability as reviewed by Kwan and colleagues41. The increase in synaptic strength lasted at least 6 days and is in agreement with frequency and amplitude increases of miniature and spontaneous synaptic currents observed in acute brain slices of mice after administering psilocybin or DMT in v/vo25,26. These drugs may increase extracellular glutamate levels, similar to ketamine, LSD and DOI administration24,50–53. To close the circle, 5-HT2A-R stimulation, subsequent glutamate release and AMPA receptor activation40 activates the BDNF-associated TrkB and mTORCI pathway which promotes BDNF expression itself54,55. Of note, PSD-95, the postsynaptic marker that we found to be upregulated, controls activity-dependent AMPA receptor incorporation in the postsynapse56,57 which can modify the strength of excitatory synaptic transmission58.

Finally, an enrichment of differentially expressed genes associated with GO terms of learning, memory and cognition suggest behavioral long-term effects of the drug which are based on the above mentioned structural and functional modifications. The effects were dependent on the duration of drug exposure suggesting that repetitive administration of psilocybin might elicit beneficial effect with respect to brain plasticity.

Together, our work confirms the postulation by the Olson group that psychedelics may act through an evolutionary conserved mechanism as we can replicate the results from animal studies in our human system25. Moreover, our study highlights the importance of a human model system for understanding the mode of action of psychedelics and potentially for disease modelling.

eLife Assessment

This fundamental study reports the effects of the psychedelic drug psilocin on iPSC-derived human cortical neurons, analyzing different aspects of structural and functional neuronal plasticity. The evidence is convincing, integrating a comprehensive characterization of 5-HT2A expression and its subcellular distribution upon treatment with psilocin at different time points. The study supports the value of using iPSC-derived human cortical neurons for testing the potentially translational effects of psilocin and other psychedelic-related compounds.

Reviewer #1 (Public review):

Summary:

This study reports the effects of psilocin on iPSC-derived human cortical neurons.

Strengths:

The characterization was comprehensive, involving immunohistochemistry of various markers, 5-HT2A receptors, BDNF, and TrkB, transcriptomics analyses, morphological determination, electrophysiology, and finally synaptic protein measurements. The results are in close agreement with prior work (PMID 29898390) on rat-cultured cortical neurons. Nevertheless, there is value in confirming those earlier findings and furthermore demonstrating the effects in human neurons, which are important for translation. The genetic, proteomics, and cell structure analyses used in this paper are its major strengths. The study supports the value of using iPSC-derived human cortical neurons for drug development involving psychedelics-related compounds.

Weaknesses:

(1) Line 140: 5-HT2A receptor expression was found via immunocytochemistry to reside in the somatodendritic and axonal compartments. However, prior work from ex vivo tissue using electron microscopy has found predominantly 5-HT2A receptor expression in the somatodendritic compartment (PMID: 12535944). Was this antibody validated to be 5-HT2A receptor-specific? Can the authors reason why the discrepancy may arise, and if the axonal expression is specific to the cultured neurons?

(2) Line 143: It would be helpful to specify the dose of psilocin tested, and describe how this dose was chosen.

(3) Figure 1: The interpretation is that the differential internalization in the axonal and somatodendritic compartments is time-dependent. However, given that only one dose is tested, it is also possible that this reflects dose dependence, with the longer time exposure leading to higher dose exposure, so these variables are related. That is, if a higher dose is given, internalization may also be observed after 10 minutes in the dendritic compartment.

(4) Figure 3 & 4: What is the 'control' here? A more appropriate control for the 24 hours after psilocin application would be 24 hours after vehicle application. Here the authors are looking at before and after, but the factor of time elapsed and perturbation via application is not controlled for.

(5) The sample size was not clearly described. In the figure legend, N = the number of neurites is provided, but it is unclear how many cells have been analyzed, and then how many of those cells belong to the same culture. These are important sample size information that should be provided. Relatedly, statistical analyses should consider that the neurites from the same cells are not independent. If the neurites indeed come from the same cells, then the sample size is much smaller and a statistical analysis considering the nested nature of the data should be used.

Reviewer #2 (Public review):

In this article, Schmidt et al use iPSC-derived human cortical neurons to test the effects the psychedelic psilocin in different models of neuroplasticity.

Using human iPSC-derived cortical neurons, the authors test the expression of 5-HT2A and subcellular distribution, as well as the effect of different times of exposure to psilocin on 5-HT2A expression. The authors evaluated the effect of the 5-HT2 antagonist ketanserin, as well as the inhibition of dynamin-dependent endocytic pathways with dynasore. Gene expression and plasticity (structural and functional) was also evaluated after different times of exposure to psilocin.

In general, results are interesting since they use the iPSC to evaluate the potentially translationally relevant effects of psilocin (the active metabolite of the psychedelic psilocybin). However, there are a few concerns that need to be addressed:

(1) My main critique is the lack of experimental validation of selectivity and/or specificity of the anti-5-HT2A antibody targeting the extracellular loop of the 5-HT2A receptor (Alomone labs, cat # ASR-033). Most of the primary antibodies targeting class A GPCRs (including the 5-HT2A receptor) have very limited selectivity. Without validation (using for example knockdown techniques to decrease expression of 5-HT2A in their iPSC-derived human cortical neurons), the experiments using this antibody should be excluded from the manuscript.

(2) Did the author evaluate whether 5-HT is present in the cell media? If it is, this may affect the functional outcomes evaluated throughout, since as the endogenous ligand it would in principle activate the 5-HT2A receptor.

(3) Some of the datasets are not statistically analyzed (or quantified), such as Figure S1F.

(4) Another important concern is the experimental design used to evaluate the effect of psilocin at different time points (24h, 4 days and 10 days). One of the unique and translationally interesting effects of psychedelics including psilocybin is that the in vivo plasticity-related effects (increased structural or synaptic plasticity for example) are observed post-acutely, or once the active compound psilocin is fully metabolized, or not present in the CNS directly targeting the 5-HT2A. Using the iPSC, it seems that the authors continuously exposed cells to psilocin (for hours or even days) at least for some of the experimental techniques. Since this is not the model of what occurs using an in vivo model (such as a single dose of psilocybin to mice, collecting frontal cortex samples 24-h after drug administration, once the active compound is fully metabolized), the authors' findings lack translational validity. Can the authors comment on this?

(5) In Figure 2E, it seems that ketamine by itself is reducing BDNF density. How then the authors conclude that ketamine blocks psi-induced effects? Using a more selective 5-HT2A antagonist such as M100907 could also improve the outcome (in terms of selectivity) of this experiment.

(6) To evaluate neurite complexity, the authors used the AAV-CamKII-mCherry viral vector, but mCherry (Fig 4A) seems to be retained in the nucleus.

(7) Minor: Reference 36- this is a review article that does not mention the psychedelic psilocin

Author response:

The following is the authors’ response to the original reviews.

Reviewer #1:

Comment 1: 5-HT2A Antibody Specificity

Was this antibody validated to be 5-HT2A receptor-specific? Can the authors reason why the discrepancy may arise, and if the axonal expression is specific to the cultured neurons?

We performed extensive validation of the anti-5-HT2A receptor antibody (Alomone #ASR-033), which is summarized in the accompanying Author response images:

Positive findings (Author response image 1c-e, Author response image 2a): (1) Western blot showed a single band at the expected molecular weight (~50 kDa) in neural progenitors and iPSCderived neurons. (2) The blocking peptide (#BLP-SR033) abolished Western blot bands and markedly reduced immunofluorescence signals in neurons, confirming epitope-specific binding.

Negative findings (Author response image 1a-b, Author response image 2a-b, Author response image 3): (1) We detected positive immunofluorescence signals in HEK293 and HeLa cells (Author response image 1a-b), which do not express 5-HT2AR. (2) Western blot also showed bands in HEK293 and HeLa cells (Author response image 2a-b). (3) Single-cell RNA-seq analysis of HEK293T cells confirmed complete absence of HTR2A expression (Author response image 3a). (4) qPCR showed no detectable HTR2A transcripts in iPSCs or HeLa cells (Ct > 36), while neural progenitors and neurons showed clear expression (Author response image 3b). (5) siRNA knockdown experiments failed to produce a corresponding decrease in immunofluorescence or Western blot signals, despite reduced HTR2A transcript levels (data not shown).

BLAST analysis: Protein BLAST analysis of the 13-amino acid immunogenic peptide sequence identified the human 5-HT2A receptor as the top hit (9/13 amino acids overlap). However, shorter sequence similarities were also found with other proteins, including APPBP1 (6/9 amino acids), Immunoglobulin Heavy Chain (6/7 amino acids), and Interleukin31 receptor (6/8 amino acids). While these partial homologies do not provide a definitive mechanistic explanation for the observed off-target binding, they illustrate that the epitope sequence is not entirely unique to the 5-HT2A receptor.

Conclusion: While our validation confirmed epitope-specific binding (blocking peptide effective in neurons), the antibody clearly detects something in cells that demonstrably lack HTR2A gene expression. This indicates off-target binding to other proteins sharing the epitope sequence. We have therefore removed all antibody-based 5-HT2A receptor experiments from the revised manuscript. This includes the receptor internalization data from Figure 1. The remaining findings (BDNF upregulation, gene expression changes, morphological effects, electrophysiology) are supported by independent methods including pharmacological blockade with ketanserin.

Comment 2: Psilocin Dose Selection

It would be helpful to specify the dose of psilocin tested, and describe how this dose was chosen.

We used 10 µM psilocin based on: (1) The seminal study by Ly et al. (2018), which demonstrated neuroplasticity effects at this concentration in rat cortical neurons. (2) Our own dose-response experiments (Figure S2B) showing maximal BDNF increase at 10 µM compared to lower concentrations (10 nM, 100 nM, 1 µM). We have clarified this in the revised Methods section.

Comment 3: Dose vs. Time Dependence

Given that only one dose is tested, it is also possible that this reflects dose dependence, with the longer time exposure leading to higher dose exposure.

We agree that dose dependence cannot be excluded with our current experimental design. This point is now moot as we have removed the 5-HT2A receptor internalization experiments from the manuscript. Future studies in our group will address dose-dependent effects on other readouts.

Comment 4: Control Conditions

What is the 'control' here? A more appropriate control would be 24 hours after vehicle application.

The control condition is indeed a vehicle (DMSO) control collected at the same time point as the experimental condition (i.e., 24 hrs post-treatment). We have clarified this in the revised figure legends and Methods section to avoid confusion.

Comment 5: Sample Size Description

The sample size was not clearly described. Statistical analyses should consider that neurites from the same cells are not independent.

We have expanded the sample size descriptions in the figure legends. Analyses were performed using 5-10 microscope images per condition, with 15 ROIs per image, across at least two independent differentiations from two genetic backgrounds. Regarding independence: each neurite segment exists within a distinct microenvironment and can be considered an independent measurement unit, consistent with established practices in the field (Paul et al., 2021, CNS Neurosci Ther). We acknowledge this increases statistical power and have noted this in the Methods.

Reviewer #2:

Comment 1: 5-HT2A Antibody Validation

Without validation (using for example knockdown techniques to decrease expression of 5HT2A), the experiments using this antibody should be excluded from the manuscript.

We agree with this assessment. As detailed in our response to Reviewer 1 (Comment 1) and documented in the Response to Reviewer Figure, our extensive validation attempts—including siRNA knockdown—could not conclusively demonstrate antibody specificity. We have removed all antibody-based 5-HT2A receptor experiments from the revised manuscript.

Comment 2: Serotonin in Cell Media

Did the authors evaluate whether 5-HT is present in the cell media?

The cell culture media used in our experiments does not contain serotonin. We have explicitly stated this in the revised Methods section.

Comment 3: Statistical Analysis of Figure S1F

Some of the datasets are not statistically analyzed, such as Figure S1F.

Figure S1F related to the 5-HT2A receptor experiments and has been removed from the revised manuscript along with the associated data.

Comment 4: Translational Validity of Prolonged Exposure

The authors continuously exposed cells to psilocin for hours or days. Since this is not the model of what occurs in vivo, the findings lack translational validity.

We acknowledge this limitation. Most experiments (BDNF, gene expression, branching) were conducted 24–48 hrs after a brief 10-minute exposure, which better reflects the in vivo situation. Prolonged exposures (96 hrs) were used specifically for synaptogenesis experiments based on literature showing that repeated LSD administration enhances spine density (Inserra et al., 2022; De Gregorio et al., 2022). Our in vitro system lacks metabolizing enzymes and glial cells, which may introduce temporal biases. We have added a discussion of these limitations in the revised manuscript.

Comment 5: Ketanserin Effect on BDNF

In Figure 2E, ketanserin by itself seems to reduce BDNF density. How do the authors conclude that ketanserin blocks psi-induced effects?

We identified that one cell line (Ctrl 1) with inherently higher BDNF density was inadvertently excluded from the ketanserin-only condition. After removing Ctrl 1 from all conditions and reanalyzing, the difference between Ctrl and Ket alone is no longer significant. The significant difference between Psi+Ket and Ket alone demonstrate that psilocin exerts effects that ketanserin can block, consistent with 5-HT2A receptor mediation. The revised figure and statistical analysis are included in the updated manuscript.

Comment 6: mCherry Localization mCherry (Fig 4A) seems to be retained in the nucleus.

The CamKII promoter drives expression of cytoplasmic mCherry, which fills the entire neuron including soma, dendrites, and axons. The apparent nuclear signal reflects mCherry accumulation in the soma, which surrounds the nucleus. The images clearly show mCherry extending into neurites, which was essential for our Sholl analysis of neuronal complexity.

Comment 7: Reference 36

Reference 36 is a review article that does not mention psilocin.

Our statement refers broadly to serotonergic psychedelics increasing neurotrophic factors. Reference 36 (Colaço et al., 2020) examines ayahuasca, which contains the serotonergic psychedelic DMT. We have revised the text to clarify this point.

Summary of Major Revisions

(1) Removed all 5-HT2A receptor antibody-based experiments from Figure 1 and supplementary figures due to inconclusive specificity validation. An Author response image documenting our validation attempts is provided.

(2) Clarified control conditions (vehicle controls at matched time points) in figure legends.

(3) Expanded sample size descriptions in Methods and figure legends.

(4) Re-analyzed ketanserin experiments with consistent cell line inclusion.

(5) Added discussion of translational limitations.

(6) Added new Figure S5 summarizing proposed signaling pathways.

(7) Expanded discussion on the relevance of iPSC-derived neurons for drug development.

Immunostaining for 5-HT2A receptor across cell types and peptide-blocking control.

(a) HEK293 cells display a positive immunofluorescent signal despite not endogenously expressing 5-HT2AR, indicating nonspecific antibody reactivity. (b) HeLa cells also exhibit a positive signal despite lacking endogenous 5-HT2AR expression, further demonstrating nonspecific antibody binding in non-expressing cell types. (c) Neural progenitor cells show clear positive 5-HT2AR staining. (d) iPSC-derived neurons exhibit robust and well-defined 5-HT2AR staining. (e) Application of the Alomone 5-HT2AR blocking peptide (#BLP-SR033) markedly reduces neuronal signal intensity, supporting epitope-specific binding.

Western blot analysis of 5-HT2A receptor abundance and peptide-blocking control.

(a-b) In line with the immunofluorescence a single band is detected in iPSCs, HEK cells, neural progenitors, iPSC-derived neurons and (b) HeLa cells. (a) Preincubation of the primary antibody with the corresponding blocking peptide abolishes this band across all samples, consistent with specific binding of the antibody to its intended epitope.

Lack of detectable 5-HT2AR expression in HEK and HeLa cells.

(a) Analysis of a human-only HEK293T single-cell RNA-seq dataset (10x Genomics; https://www.10xgenomics.com/datasets/293-t-cells-1-standard-1-1-0, accessed 2025-11-25) shows no meaningful HTR2A expression, whereas other genes such as GAPDH, TP53, MYC, and ACTB are robustly detected. Consistently, evaluation of a “Barnyard” dataset - an equal mixture of human HEK293T and mouse NIH3T3 cells (10x Genomics; https://www.10xgenomics.com/datasets/20-k-1-1mixture-of-human-hek-293-t-and-mouse-nih-3-t-3-cells-3-ht-v-3-1-3-1-high-6-1-0, accessed 2025-1125) reveals only ~4 of ~10,000 droplets with minimal HTR2A signal, confirming the absence of meaningful expression.(b) (b) qPCR analysis further demonstrates no detectable HTR2A transcripts in iPSCs or HeLa cells (Ct > 36), while neural progenitors and iPSC-derived cortical neurons show expression when normalized to housekeeping genes GAPDH and TBP.

Version 2

Abstract

Psilocybin is studied as innovative medication in anxiety, substance abuse and treatment-resistant depression. Animal studies show that psychedelics promote neuronal plasticity by strengthening synaptic responses and protein synthesis. However, the exact molecular and cellular changes induced by psilocybin in the human brain are not known. Here, we treated human cortical neurons derived from induced pluripotent stem cells with the 5-HT2A receptor agonist psilocin - the psychoactive metabolite of psilocybin. We analyzed how exposure to psilocin affects gene expression, neuronal morphology, synaptic markers and neuronal function. Psilocin provoked a 5-HT2A-R-mediated augmentation of BDNF abundance. Transcriptomic profiling identified gene expression signatures priming neurons to neuroplasticity. On a morphological level, psilocin induced enhanced neuronal complexity and increased expression of synaptic proteins, in particular in the postsynaptic-compartment. Consistently, we observed an increased excitability and enhanced synaptic network activity in neurons treated with psilocin. In conclusion, exposure of human neurons to psilocin might induce a state of enhanced neuronal plasticity which could explain why psilocin is beneficial in the treatment of neuropsychiatric disorders where synaptic dysfunctions are discussed.

Introduction

Since the 90’s the mind-altering psychedelics have become again the subject of an intriguing stream of research in psychiatric disorders1,2. Particularly, in times where the development and the approval of new medications is decreasing and the number of patients suffering from psychiatric disorders is rising there is a huge need for new therapeutical interventions3,4. Compared to classical drugs the broad therapeutical effect of psychedelics is rapid, robust and can be long-lasting even after single administration5. In particular, the 5-hydroxytryptamine receptor 2A (5-HT2A-R) targeting psilocybin, the compound of the so-called “magic mushrooms”, is discussed as fast-acting and long-lasting antidepressant in treatment-resistant depression (TRD), anxiety, obsessive-compulsive disorder and addiction6–11. Not surprisingly psilocybin is stated as “breakthrough therapy” by the United States Food and Drug Administration (FDA) in the treatment of depression since 2019. While we know that 5-HT2A- R activation plays a principal role in serotonergic psychedelic-mediated behavioral and cellular response12–22 the molecular and cellular changes induced in the brain - on a single cell and network level are barely understood. The recent demonstration of intracellular 5-HT2A-Rs additionally increased the complexity of psychedelic actions23. The administration of psychedelics may enable brain network resetting1 by generating a plastic cellular state in which synaptic remodeling and augmentation of neuroplasticity-associated proteins and genes are likely24. Indeed, biological evidence for the “resetting” hypothesis comes from a pioneering study by Olson and colleagues in 2018 which showed that the treatment with serotonergic psychedelics increases the synthesis of synaptic proteins, strengthens synaptic responses, and fosters neurito- and synaptogenesis in rat cortical neurons25. The group therefore introduced the term “psychoplastogen” (greek: psych- (mind), - plast (molded), - gen (producing) for underlining their plasticity-promoting properties. Moreover, serotonergic psychedelics already have been shown to promote the 5-HT2A-R-mediated growth of dendritic spines and modulate neurotransmission25–27. As molecular and cellular psychedelic research is recently based nearly exclusively on animal studies, the question emerged whether those insights can be translated to the human brain. Psychiatric disorders and psychedelic effects are complex, multisymptomatic and therefore often difficult to study in non-human model organisms. Most importantly, drugs that are efficiently tested in psychiatric animal models might not be necessarily transferable to the human system28. In that context, induced human pluripotent stem cells (iPSC) have emerged as a powerful tool to generate neurons and neuronal circuitries, model brain disorders and identify the molecular mechanisms of drug interventions29.

Here, we explored the molecular, transcriptional, morphometric and functional consequences of the psychoactive 5-HT2A receptor agonist psilocin, the active metabolite of psilocybin in human iPSC- derived cortical neurons. We demonstrate that psilocin leads to a set of molecular, morphological and functional changes that start shortly after administration and manifest in time. These include an increase in brain derived neurotrophic factor (BDNF) expression and an activation of gene expression programs associated with neuromodulation and plasticity resulting in increased neuronal complexity, synaptogenesis and changes in neuronal network function. Thus, our study provides first evidence that the 5-HT2A receptor agonist psilocybin activates widespread neuroplastic programs in human neurons.

Results

Differentiation of human iPSCs in human cortical neurons

As an experimental model to study the effects of psilocin in human neurons and neuronal networks we differentiated human iPSCs (expressing the pluripotency-associated transcription factors OCT4 and SOX2) into neural progenitor cells which we further differentiated into neurons of mainly dorsal forebrain identity (Fig. 1A and Fig. S1A-B)30–33. Neural progenitor cells express typical neural stem / progenitor cell markers such as NESTIN or SOX2 as well as the dorsal forebrain-associated transcription factors PAX6 and are negative for FOXA2, as floorplate/ midbrain marker (Fig. S1B). Neurons differentiated from these progenitors for >5 weeks express the neuronal antigen NeuN, the dendritic marker MAP2, and the axonal marker TAU. Most neurons express TBR1 and / or CTIP2, transcription factors typically found in Layer 5/6 neurons of the human cortex (Fig. 1A). Neuronal activity in mature neurons is indicated by expression of the activity-dependent immediate early gene cFOS (Fig. 1A). Astrocytes expressing glial fibrillary acid protein (GFAP) were found occasionally (Fig. 1A). These represent the only contamination in the otherwise purely neuronal culture, which contains no residual progenitors (Fig. S1C). 5-HT2A-R expression was confirmed by PCR (Fig. S1D)34,35.

Validation of mature cortical neuron properties.

(A) hiPSCs were differentiated in vitro into glutamatergic cortical neurons over a neuronal progenitor step since the cerebral cortex is a key region for psychedelic effects and psychiatric disorders. Mature 40-day old cortical neurons were glutamatergic (vGLUT2 expression) and expressed cortical layer markers like TBR1, CTIP2 and the neuronal markers NeuN, MAP2 and TAU. cFOS expression is a sign for neuronal activity. GFAP expression indicated a low amount of astrocytes. Scale bar: 50 pm, for GFAP staining 100 pm.

Psilocin induces BDNF expression and downstream signaling in human neurons

Psychedelics have been shown to efficiently increase levels of neurotrophic factors, such as BDNF12’25’36’37. We thus addressed BDNF expression in human neurons following exposure to psilocin. We analyzed the effect of different psilocin concentrations (ranging from 10 nM to 10pM) exposed for 10 min and also at 24 hrs following exposure. We observed a dosage-dependent increase in the abundance of BDNF-positive particles (quantified as particles per pm neurite length) which reached significance at a concentration of 10 μM psilocin (Fig. S2A-B). A longer treatment (6 hrs instead of 10 min) with a lower concentration (100 nM instead of 10 pM) of psilocin also induced BDNF abundance significantly, but less strikingly (Fig. S2C-D). We thus continued to use 10 μM psilocin in the following experiments. The clear induction of BDNF particle density was reproduced in four biological independent control cell lines and several differentiation batches for each line (from MCtrl = 1.5 ± 1.2; MPsi = 2.6 ± 1.9) (Fig. 2A-C). When comparing BDNF densities independently in dendrites (MAP2-positive) and axons (Tau-positive), the significant increase in BDNF could be assigned to both neuronal compartments to a comparable extent (Fig. S2E-F). Also, at 48 hrs following the exposure to psilocin, a significant increase of BDNF particles was observed compared to the untreated condition (Fig. S2G). The increase in BDNF abundance was 5-HT2A-R-mediated as treatment of the neurons with the specific 5-HTR2A-R antagonist ketanserin prevented BDNF upregulation by psilocin (Fig. 2D-E) (from MCtrl = 1.8 ± 0.8; MPsi = 3.4 ± 1.7; MPsi + Ket = 1.5 ± 0.7; MKet = 1.4 ± 1.2). Upregulation of BDNF could also be prevented by blocking CME with dynasore or protein kinase C (PKC) with the PKC inhibitor chelerythrine, indicating that CME and PKC activation are critically involved in this process (Fig. 2F-G) (from MCtrl= 3.1 ± 1.5 to MPsi = 4.6 ± 2.4; MPsi + Chel = 3 ± 1.7; MPsi + D = 3.4 ± 1). Upregulation of BDNF was also validated by Western immunoblot where we observed an increase in the levels of pro- BDNF and were able to detect mature BDNF at approximately 14 kDa, which, under baseline conditions, was below the detection level (Fig. S2H). Increased BDNF signaling in psilocin-treated neurons is indicated by an increase in the phosphorylation of the BDNF receptor TrkB (Fig. S2H). Furthermore, as a downstream target of BDNF signaling we observed an increase of the phosphorylation of AKT at Ser 473 1 day (Fig. S2I) and 3 days after psilocin exposure (Fig. 2H-I) which was reversed by ketanserin treatment (from MCtrl = 0.8 ± 0.2; MPsi = 1.3 ± 0.2; MKet = 0.9 ± 0.2).

Psilocin-induced increase in BDNF level was 5-HT2A receptor and PKC- and endocytosis mediated and induced activation of m-BDNF/TrkB-associated downstream pathway.

(A-B) Representative image of a neuronal network for pre-treatment condition (Ctrl) and 24 hrs after a 10 min 10 μM short psilocin trigger (Psi). Scale bar: 50 pm, (B) close up: 2 pm. (C) BDNF density was significantly increased 24 hrs after a 10 min 10 μM psilocin trigger (Psi). Four Ctrl cell lines were included in the analysis (Ctrl with N = 613 neurites; Psi with N = 529 neurites). (D) Representative image 24 hrs after the simultaneous treatment with psilocin and ketanserin (Psi + Ket) and single treatment with ketanserin (Ket), scale bar: 50 pm. (E) Ketanserin co-treatment (Psi + Ket) and ketanserin monotreatment (Ket) significantly reduced BDNF density compared to the 24 hrs monotreatment psilocin condition, suggesting a 5-HT2A-mediated process. Ketanserin monotreatment provoked a significant reduction in BDNF density compared to ketanserin co-treatment with psilocin. Two Ctrl cell lines were included in the analysis, each with one biological batch (Ctrl with N = 99 neurites; Psi with N = 102 neurites; Psi + Ket with N = 102 neurites; Ket with N = 102 neurites). (F) Representative image 24 hrs after the simultaneous treatment of 10 μM psilocin with chelerythrine (Psi + Chel) or dynasore (Psi + D), scale bar: 50 pm. (G) Chelerythrine (Psi + Chel, selective PKC inhibitor) or dynasore (Psi + D, inhibition of clathrin-coated vesicle invagination) co-treatment significantly reduced BDNF density compared to the psilocin monotreatment condition (Psi). Two Ctrl cell lines were included in the analysis, each with one biological batch (Ctrl with N = 90 neurites; Psi with N = 90 neurites; Psi + Chel with N = 90; Psi + D with N = 90 neurites). (H) Phosphorylated AKT (pAKT) protein level was increased 72 hrs after psilocin exposure (Psi), reversed by ketanserin co-treatment (Psi + Ket), (I) both effects were not significant. Ctrl cell line 3 was included in the analysis (Ctrl with N = 4 data points, Psi with N = 4 data points, Psi + Ket with N = 4 data points), each with two biological batches. For all analyses Kruskal-Wallis-Test for independent samples was calculated. Post hoc Wilcoxon rank sum test. Bonferroni-correction, adjusted p < .05, mean ± SD. Significance levels against the respective control and for multiple group comparisons are *p < .05.

Psilocin induces gene expression changes associated with axonal and synaptic plasticity

To analyze the influence of psilocin on global gene expression in cortical neurons, we performed whole transcriptome sequencing one day and three days following a single 10 min administration with 10 μM psilocin in neurons from two independent genetic backgrounds. GO enrichment and KEGG pathway analysis comparing cells one day and three days after psilocin administration with the respective controls revealed an enrichment of significantly affected genes in many ontologies and pathways associated with axonal growth and synaptic remodeling, plasticity and learning, memory and cognition (Fig. 3A, S3A). Most significant changes occurred within the first 24 hrs which is reflected when plotting the significantly changed genes only for selected GO term enriched at both time points (Fig. 3B, S3B). When looking at a more global gene expression level of all genes included in the respective GOs, gene expression of GOs associated with axonal outgrowth show a generally higher expression at day 1 which is further increased at day 3. In contrast genes associated to synaptic organization, plasticity and learning, memory and cognition show a trend towards decreased expression at day 1 and a generally stronger increase at day 3 (Fig. 3C, S3C). The overlapping assignment of significant differentially expressed genes to the aforementioned GOs underlines their importance in long-term plasticity (e.g., CDK5, CAMK), synapse formation (e.g., SYN1) and axonal and neurite growth (e.g., GAP43) and neuronal structure marker (e.g., MAPT) (Fig. 3D). The latter is consistent with increasing staining intensity for TAU (Fig. 2). We further found a strong significant upregulation of immediate early genes (lEGs) (e.g., FOSL2, JUND, EGR1, ARC, FOSB) and of the excitatory glutamatergic AMPA/NMDA receptor genes (GRIA/GRIN) after psilocin administration (Fig. 3E). One day but not three days of psilocin treatment more strongly upregulated the AMPA genes GRIA1, 2, 3 than the NMDA genes GRIN1, 2B, 2C (Fig. 3E), supporting modulation of glutamatergic excitatory signalling genes upon psilocin treatment. These effects could be reversed upon co-treatment with antagonist ketanserin, indicating that most changes are 5-HT2A receptor mediated (Fig. 3F).

Psilocin displays fast and enduring changes of the genetic landscape.

(A) Enrichment in differentially expressed genes associated with synapse formation, neuronal plasticity and axonogenesis GO terms 1 day after psilocin administration, and the effect 3 days later. (B) Psilocin induced within 24 hours a first wave of upregulation of selected GO genes based on DESeq2 normalized counts. Log2fold changes of each differentially expressed genes between two conditions are shown as dots. Zscores indicate if more genes in the respective GO term are upregulated or downregulated also indicated by height and color of the bars (C) TPM-normalized mean (red line) of psilocin-induced temporal expression pattern of genes belonging to the indicated GO. Shaded area is indicating SD. (D) Chord plots showing significant genes appearing in at least 4 GO terms (of 6 selected GO terms) 1 day after psilocin administration and appearing in at least 3 GO terms 3 days after psilocin administration. (E). Heat map (z-scaled normalized counts) showing expression of immediate early genes and AMPA/ NMDA receptor genes throughout psilocin administration. (F) Differentially expressed genes, that are up-/ downregulated upon psilocin treatment, showed a reversed effect upon co-treatment with ketanserin. Mean TPM-normalized expression values are shown (red line), shaded area indicates S.D. BP, biological process; CC, Cellular Compartment; KEGG, Kyoto Encyclopedia of Genes and Genomes. TPM, transcripts per kilobase million. Significance levels against the respective control ns: p-adj. > .05, *: p-adj. <= .05, **: p-adj. <= .01, ***: p-adj. <= .001.

Psilocin increases neurite complexity

The gene expression analysis indicated an induction of morphometric changes by psilocin in particular on neurites. Therefore, we assessed neuronal complexity performing Sholl analysis in neurons differentiated for >40 days at 24 hrs and 48 hrs following psilocin exposure. To identify the dendritic arbor of single mature neurons in these cultures, the cultures were transduced with adeno-associated vectors (AAVs) coding for mCherry under control of the CaMKIIa promoter (Fig. 4A-D).

Psilocin induced neurite branching.

(A-C) Cells were transduced with AAV CamKIIa p-hCHR2(134a)-mCherry. (A) Representative image for pre-treatment condition (Ctrl) and 48 hrs after a 10 min short psilocin trigger (48 hrs) for mCherry staining showed an increase in neurite intersections for the latter. (B) Singular analysis: 48 hrs after a 10 min 10 μM psilocin trigger the number of primary (25 μm distance from soma) neurite intersections significantly increased compared to the untreated control condition (Ctrl). The number of intersections at 50 μm distance from soma significantly increased 24 hrs and 48 hrs after a 10 min 10 μM psilocin trigger. (C) Significant changes for the calculated total neurite length and for the total number of neurite intersections significantly increased 24 hrs and 48 hrs after a 10 min 10 μM psilocin trigger. Two Control cell lines with two biological batches (Ctrl with N = 43 - 48 neurites; 24 hrs with N = 47 - 48 neurites; 48 hrs with N = 31 - 35 neurites) were included. (D) Sholl analyses summary for results shown in figure (B). For all analyses Kruskal-Wallis-Test for independent samples was calculated. Post hoc Wilcoxon rank sum test. Bonferroni-correction, adjusted p < .05, mean ± SD. Significance levels against the respective control are *p < .05.

We observed an increase of primary neurites at 25 μm (significant at 48 hrs and a trend at 24 hrs from MCtrl = 4.1 ± 2.1; M24hrs = 5.4 ± 3.7; M48hrs = 5.5 ± 2.4) and of intersections at 50 μm (significantly increased at 24 and 48 hrs post treatment from MCtrl = 2.7 ± 1.4; M24hrs = 4 ± 2.3 to M48hrs = 3.9 ± 1.8; Fig. 4B). As a result, the calculated total neurite length was significantly increased at 24 and 48 hrs post treatment from MCtrl = 272.5 ± 142.7; M24hrs = 383.2 ± 209.9; M48hrs = 352.8 ± 162.5 (Fig. 4C). And we observed a significant increase in the total number of intersections in psilocin-exposed neurons quantified in steps of 25 μm up to a distance of 125 μm (MCtrl = 10.9 ± 5.7; M24hrs = 15.3 ± 8.4, M48hrs = 14.1 ± 6.5; Fig. 4D). These data indicate, that gene expression changes elicited by psilocin resulted in neurite outgrowth and an increase of global dendritic complexity as early as 24h following a single 10 min exposure.

Psilocin increases synaptic strength and synaptogenesis

To address changes on neuronal function, we exposed neuronal cultures 10 min or 24 h to 10 μM psilocin and performed whole-cell patch-clamp experiments one week later. Following the 10 min exposure, we observed a increasing trend in the number of evoked action potentials (APs), having amplitudes of around 100 mV (Fig. 5A). Extended exposure of the cultures to psilocin (24h) increased this effect reaching significance (total number of APs from MCtrl = 20.1 ± 12; MPsi 10min; day 7 = 28.9 ± 16.9; MPsi 24hrs; day 7 = 30 ± 14.5; AP amplitude from MCtrl = 101.1 ± 13.3; MPsi 10min; day 7 = 101.1 ± 13.2; MPsi 24hrs; day 7 = 101.3 ± 16.7). To find out whether the increase in AP firing enhanced synaptic network activity, we recorded spontaneous AP-dependent excitatory postsynaptic currents (sEPSCs). Indeed, the frequency of sEPSCs increased in both psilocin-treated conditions to some extent (MCtrl = 0.6 ± 0.7; MPsi 10min; day 7 = 1.0 ± 1.3; MPsi 24hrs; day 7 = 1.1 ± 1.2; Fig. 5B). And, we observed a slight increase in sEPSC amplitude (MCtrl = 14.2 ± 4.7; MPsi 10min; day 7 = 16.7 ± 5.1; MPsi 24hrs; day 7 = 15.5 ± 5.3; Fig. 5B), consistent with the observed GRIA upregulation (Fig. 3E). The latter amplitude increase was next examined on AP-independent miniature EPSCs (mEPSCs) which are generally caused by the spontaneous release of single vesicles (Figs. 5C, S4A). The amplitudes of mEPSCs increased 7 days after the start of 24 hrs psilocin administration consistently for two cell lines (MCtrl = 11.1 ± 2.2; MPsi 24hrs; day 7 = 12.8 ± 4.2, Fig. 5C; MCtrl = 9.3 ± 2.8; MPsi 24hrs; day 7 = 11.2 ± 3.7, Fig. S4A), whereas the frequencies of the mEPSCs were rather stable (MCtrl = 0.8 ± 0.5; MPsi 24hrs; day 7 = 0.9 ± 0.6, Fig. 5C; MCtrl = 0.9 ± 0.7; MPsi 24hrs; day 7 = 0.7 ± 0.5, Fig. S4A). When examining mEPSCs from separately differentiated neurons exposed to psilocin for 24 hrs and analyzed at day 11, the increase in mEPSC amplitude was somewhat attenuated compared with 7 days (MCtrl = 11.7 ± 4.3; MPsi 24hrs; day 11 = 12.4 ± 3.6; frequency from MCtrl = 0.8 ± 0.8; MPsi 24hrs; day 11 = 0.5 ± 0.3, Fig. 5C). Repeated LSD administration altered gene and protein expression related to neuroplasticity signaling in the mouse prefrontal cortex and increased dendritic spine density38,39. We therefore hypothesized that extended psilocin administration fosters synaptic network activity and synaptogenesis and thus examined effects of a 96 hrs exposure. In line with the above-mentioned hypothesis, the amplitudes of the sEPSCs increased significantly in two lines (MCtrl = 22.5 ± 6.5; MPsi 96hrs; day 10 = 30.7 ± 4.8, Fig. 5D; MCtrl = 15.9 ± 4.5 ; MPsi 96hrs; day 10 = 27.6 ± 10, Fig. S4B), and the frequencies showed a trend to increase similar to Fig. 5B, reflecting enhanced network activity (MCtrl = 1.0 ± 1.0; MPsi 96hrs; day 10 = 1.5 ± 1.0, Fig. 5E; MCtrl = 1.5 ± 1.7; MPsi 96hrs; day 10 = 1.0 ± 0.7, Fig. S4B).

Psilocin-induced increase in synaptic strength and synaptogenesis.

(A) Total number of evoked action potentials (eAPs) significantly increased at day 7 after 24hrs permanent psilocin administration (Psi 24hrs; day 7) and was increased 7 days after a 10 minutes psilocin trigger (Psi 10min; day 7). AP amplitudes stayed constant. Representative traces for total number of APs. One Ctrl cell line with 2 biological batches was included in the analysis. Ctrl 3 with N = 28 cells, Psi 10min; day 7 with N = 29 cells, Psi 24hrs; day 7 with N = 34 cells. (B) Increase in sEPSCs amplitude and frequency 7 days after 10 minutes and 24hrs permanent psilocin administration. One Ctrl cell line with 2 biological batches was included in the analysis, Ctrl 3 with N = 19 cells, Psi 10min; day 7 with N = 20 cells, Psi 24hrs; day 7 with N = 21 cells. (C) Increase in mEPSCs amplitude 6 days after permanent 10 μM psilocin administration, Ctrl 3 with N = 16 cells, Psi 24hrs; day 7 with N = 14 cells. Increase in mEPSCs amplitude 10 days after 24hrs permanent psilocin administration, Ctrl 3 with N = 15 cells, Psi 24hrs; day 11 with N = 20 cells. (D) Significant increase in sEPSCs amplitude 6 days after 96hrs permanent 10 μM psilocin administration (Psi 96hrs; day 10). Ctrl 3 with N = 18 cells, Psi 96hrs; day 10 with N = 13 cells. Representative traces for control and psilocin condition. For all experiments Mann-Whitney-U-test for independent samples was calculated, mean ± SEM. Significance levels against the respective control are *p < .05. (E-F) Representative dendritic PSD-95 and synapsin staining for the untreated condition (Ctrl), 4 days after permanent 10 μM psilocin administration (4 days) and 10 days after 4 days permanent 10 μM psilocin administration (10 days). (E) Scale bar: 50 μm (F), closeup: 10 pm. Trend in synapsin (G) density and (H) intensity increase 4 days after permanent psilocin administration. (I) PSD-95 particle per neurite length and (J) intensity per area and (K) synapsin/ PSD- 95 co-localization were significantly increased 4 and 10 days after 4 days permanent 10 μM psilocin treatment compared to an untreated control condition. (G-K) Ctrl with N = 180 neurites, 4 days with N = 180 neurites, 10 days with N = 180 neurites. Two control cell lines each with two biological batches were included. For all analyses Kruskal-Wallis-Test for independent samples was calculated. Bonferroni- correction, adjusted p < .05, mean ± SD. Significance levels against the respective control are *p < .05.

This enhanced synaptic network activity should affect the abundance of either the presynaptic marker synapsin and/or the postsynaptic marker PSD-95 and may even affect their co-localization at day 4 and at day 10 following 4 day permanent 10 μM psilocin administration. In line with above electrophysiological results for the 10 day psilocin exposure (Fig. 5D) and the gene expression data, which indicated changes associated with synaptogenesis and synaptic plasticity (Fig. 3) we observed for the pre-synaptic marker synapsin, a trend towards an increase of the density (particles per neurite length) (from MCtrl = 0.25 ± 0.2; M4days = 0.28 ± 0.2; M10days = 0.29 ± 0.2) and intensity (mean fluorescence intensity of the particles) (from MCtrl = 21 ± 13.3; M4days = 24.1± 15; M10days = 19.8 ±10.9) at both time points (Fig. 5E-H). In line with the gene expression data (Fig. 3A, postsynaptic density), we observed a significant increase in the density of the postsynaptic marker PSD-95 (from MCtrl = 1.1 ± 0.5; M4days = 1.4 ± 0.6; M10days = 1.3 ± 0.5) and in its intensity (from MCtrl = 31.8 ± 12,9; M4days = 36.8 ± 15.6; M10days = 38 ± 16.4) for both time points (Fig. 5I-J). These effects were accompanied by a significant increase in the co-localization of both markers (from MCtrl = 0.2 ± 0.2; M4days = 0.3 ± 0.2; M10days = 0.3 ± 0.2; Fig. 5K). Together, these experiments indicate that psilocin augments synaptic strength primarily via an increase in the postsynaptic receptor density, which lasts at least for 6 days from the start of psilocin withdrawal and that extended exposure of neurons to psilocin pronounces the effects on synaptic strength.

Discussion

The neuroplasticity-promoting psychoplastogen25 psilocybin is currently developed as a new medication in the treatment of psychiatric disorders1,2,40. “Brain network resetting”1 by psilocybin , i.e. the restoration of neuronal and synaptic dysfunction associated with the pathophysiology of mental disorders41–45, could be achieved by stimulating signaling pathways associated with neuroplasticity and BDNF signaling24,41. So far, experimental evidence for this hypothesis is based on data from non-human model organisms.

In our experimental setting, we observed a pronounced psilocin-induced BDNF upregulation in human cortical neurons and show that PKC activation and endocytosis, two mechanisms contributing to 5- HT2A-R internalization42,43, are involved as chelerythrine and dynasore inhibited BDNF augmentation. BDNF plays an important role in neurogenesis, synaptogenesis, and the formation of synaptic interactions the basis of neuroplasticity (Fig. S5), the brain’s enduring ability to constantly change in response to influences and experiences and lifelong learning44. Particularly the mature BDNF/phosphorylated-TrkB receptor complex is involved in multiple pathways linked to prosurvival and synaptic plasticity44. Recently it was shown that psychedelics also directly bind to the BDNF TrKB receptor, thereby affecting TrkB dimerization and facilitating the effect of endogenous BDNF, which underlines the importance of the BDNF system for the action of psychedelics45. In keeping with this, we show that psilocin leads to a fast and enduring upregulation of proteins and genes linked to neuronal complexity, synaptogenesis and synaptic transmission, as also demonstrated in psychedelic-mediated plasticity in animal studies25,26,46. Of note, BDNF upregulation and changes in gene expression were reversed by ketanserin confirming that these effects are mediated by 5-HT2A-Rs.

In our model, augmentation of neuronal complexity is an outcome which can be detected as early as 24 hrs after psilocin administration. In rodents, serotonergic psychedelics also foster synaptogenesis and spinogenesis25,26,46. In line with this we show that psilocin also promotes synaptogenesis, measured by the increase in PSD-95, synapsin and their colocalization. As a result, we observe changes in intrinsic neuronal properties and network function. More specifically, we see an increased excitability and an increase in postsynaptic current frequency but in particular amplitude. The higher number of action potentials generated by current injections could be due to increased dendritic excitability as reviewed by Kwan and colleagues41. The increase in synaptic strength lasted at least 6 days and is in agreement with frequency and amplitude increases of miniature and spontaneous synaptic currents observed in acute brain slices of mice after administering psilocybin or DMT in vivo25,26. These drugs may increase extracellular glutamate levels, similar to ketamine, LSD and DOI administration24,47–50. To close the circle, 5-HT2A-R stimulation, subsequent glutamate release and AMPA receptor activation40 activates the BDNF-associated TrkB and mTORC1 pathway which promotes BDNF expression itself51,52. Of note, PSD-95, the postsynaptic marker that we found to be upregulated, controls activity-dependent AMPA receptor incorporation in the postsynapse53,54 which can modify the strength of excitatory synaptic transmission55.

Finally, an enrichment of differentially expressed genes associated with GO terms of learning, memory and cognition suggest behavioral long-term effects of the drug which are based on the above mentioned structural and functional modifications. The effects were dependent on the duration of drug exposure suggesting that repetitive administration of psilocybin might elicit beneficial effect with respect to brain plasticity.

Our study has some limitations. The in vitro system lacks metabolizing enzymes and the cellular complexity of brain tissue including glial cells. Furthermore, some experiments required prolonged psilocin exposure (96 hrs), which does not fully recapitulate the in vivo situation where psilocin is rapidly metabolized. Nevertheless, the majority of our key findings (BDNF upregulation, gene expression changes, neurite branching) were observed 24-48 hrs after a brief 10-minute exposure, more closely reflecting the post-acute effects observed in vivo.

Together, our work confirms the postulation by the Olson group that psychedelics may act through an evolutionary conserved mechanism as we can replicate the results from animal studies in our human system25. Our study underscores the significance of human-based models, particularly iPSC-derived cortical neurons, in elucidating the mechanisms of psychedelics and their potential for disease modeling and drug development. By leveraging these human-derived systems, the research emphasizes the translational potential of psychedelics, offering deeper insights into their action in a human context and advancing therapeutic innovation56.